---
title: "dermacentor"
author: "Jessica Martin, Ilya Fischhoff"
date: "1/23/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#####install packages
```{r packages, echo=FALSE}
pkgTest <- function(x)
{
  if (x %in% rownames(installed.packages()) == FALSE) {
    install.packages(x, dependencies= TRUE)    
  }
  library(x, character.only = TRUE)
}
neededPackages <- c("gbm", "caTools", "dismo", "caret", "ROCR", "ggplot2")

for (package in neededPackages){pkgTest(package)}
```


###read in data and rearrange columns
```{r read}
tick2<-read.csv("Dermacentor database_v2_01-10-20_4R.csv", header = T)

#remove variables with all NAs
tick <- tick2[, !apply(is.na(tick2),2,all)]

#rearrange columns so that GIDEON (the response variable) is last
tick<-tick[c(1:91,93:101,92)]

```

#### remove columns with near zero variance Global
### this reduces the number of variables from 101 to 92
```{r near zero}
nzv <- nearZeroVar(tick,saveMetrics=TRUE, freqCut = 80/20)#changed from default 95/5 because otherwise ran into checksum error when doing bootstrap AUC
nzv <- row.names(nzv[which(nzv$nzv==TRUE),])
dropnzv<-names(tick[ , which(names(tick) %in% nzv)])
tick<-tick[ , -which(names(tick) %in% nzv)]
dim(tick)
```

### select traits above a coverage threshold

## Ilya - Barbara's orignal comment here said to "update to inlcude the residuals columns"
## I'm not sure that I did this step right (I'm unclear on what the residuals columns are for)
## Does this step make sense to you?

```{r}
index<-c(2:dim(tick)[2]) #comment from Jess: index excludes columns for species.name, does not include residuals (?)

traits<-names(tick)[index]
output<-c() # trait coverage

for(i in 1:length(traits)){
  output[i]<-length(tick[which(is.na(tick[,index[i]])==F),index[i]])/length(tick[,index[i]])
}
coverage<-data.frame(cbind(traits,output),stringsAsFactors = FALSE)
colnames(coverage)<-c("Predictor","Coverage")

# set coverage threshold
#JM: changed threshold from 0.01 to 0.1
traits<-coverage[which(coverage$Coverage > 0.1),]  

#remove GIDEON
traits = subset(traits, Predictor != "GIDEON")
```


##run GBM
```{r gbm}
shrinkage = 0.0001
n.trees = 80000
interaction.depth = 3
n.minobsinnode = 2
cv.folds = 4

#Start the clock
ptm<-proc.time()

set.seed(777)
intrain<-createDataPartition(y=tick$GIDEON,
                             # groups=3,
                             p=0.7,
                             list=FALSE)
train<-tick[intrain,]
test<-tick[-intrain,]


# column 94 (GIDEON) contains the outcomes that the model is trying to predict
y_col = which(names(tick)=="GIDEON")
model<-as.formula(paste(colnames(tick)[y_col], "~",
                        paste(traits$Predictor,collapse = "+"), #traits
                        collapse="+"))
save(model, file = "model.Rdata")
tickgbm <- gbm(model,
               data=train, 
               distribution="bernoulli", 
               n.trees=n.trees, 
               shrinkage=shrinkage,
               interaction.depth=interaction.depth,
               bag.fraction=0.50,
               train.fraction=1,
               n.minobsinnode=n.minobsinnode,#too few data for 3
               cv.folds=cv.folds,#too few data for cv.folds = 5
               class.stratify.cv = TRUE,
               keep.data=TRUE,
               verbose=TRUE,
               n.cores=1)

#check performance
best.iter <- gbm.perf(tickgbm,method="cv",plot.it=TRUE) #this gives you the optimal number of trees based on cv performance, other methods will over or under predict
print(best.iter)

gbmsum<-summary(tickgbm, n.trees=best.iter, method=relative.influence)
gbmsum

# # predictions on the TRAINING SET
output<-predict(tickgbm, newdata=train, n.trees=best.iter, type="response") 
output<-cbind(output,as.numeric(train$GIDEON))
colnames(output)<-c("output","label")
output<-output[order(-as.numeric(output[,1])),]
plot(output)

# # training AUC for Bernoulli distributed responses
auc=colAUC(output[,1],output[,2])
print(auc)
pred<-prediction(output[,1],output[,2])
perf<-performance(pred,"tpr","fpr")
plot(perf,colorize=TRUE,main="ROC full model")

# Predictions on the TEST set
output.test<-predict(tickgbm, newdata=test, n.trees=best.iter, type="response") 
output.test<-cbind(output.test,as.numeric(test$GIDEON))
colnames(output.test)<-c("output","label")
output.test<-output.test[order(-output.test[,1]),]
plot(output.test)

## test AUC for Bernoulli distributed responses
auctest=colAUC(output.test[,1],output.test[,2])
print(auctest)
predtest<-prediction(output.test[,1],output.test[,2])
perftest<-performance(predtest,"tpr","fpr")
plot(perftest,colorize=TRUE,main="ROC full model",add=T)

#save gbm model output
save(tickgbm, file = "tickgbm.Rdata")

#Stop the clock
(proc.time()-ptm)/60

```

##make plot of relative influence
```{r}
load("tickgbm.Rdata")
 x = summary(tickgbm)
# 
 x.df= data.frame(variable = x$var,
                  relative.influence = x$rel.inf)
x.df$variable=as.character(x.df$variable)

x.df = subset(x.df, relative.influence >= 1)
x.df$variable = factor(x.df$variable, levels = x.df$variable[order(x.df$relative.influence)])
save(x.df, file = "x.df.Rdata")
plot = ggplot(data = x.df, aes(x = variable, y =relative.influence))+
  ylab("relative influence (%)")+
  xlab("variable")+
  geom_bar(stat="identity")+
  coord_flip()
# 
ggsave(plot = plot, filename = "Figure.relative.influence.jpg")


```

##Bootstrap permutations for null distribution of AUC -- presence/absence
```{r boot_null_AUC}
  # set.seed(Sys.time())#i think we need to reset seed

p = 5#number of bootstrap permutations
# p = 1#number of bootstrap permutations
# save(p, file = "p.Rdata")
df = tick
rm = "species.name"#remove this field
keep = setdiff(names(df), rm)
df = df[,keep]
#Start the clock
ptm<-proc.time()

permutedAUC<-c()
best.iter2.list = c()
bootstrap_runs = p
i=1
while (i <= bootstrap_runs) {
  # for permutation loop
  ## random permutation of Label
  randomLabel<-sample(df$GIDEON)

  pan2<-cbind(randomLabel,df)
  #remove previous label
  rm = "GIDEON"
  keep = setdiff(names(pan2),rm)
  pan2 = pan2[,keep]

  pan2[,1]<-sapply(pan2[,1],as.character)

  ## create training and test sets
  intrain2<-createDataPartition(y=pan2$randomLabel,
                                p=0.7,
                                list=FALSE)

  test2<-pan2[-intrain2,]
  training2<-pan2[intrain2,]

  check<-1-is.na(training2)*1
  checksum<-apply(check,2,sum)
  n_cols= dim(training2)[2]

  
  if(length(which(checksum>=2))==n_cols){#this makes sure we don't get any columns with all zeros. Should be == to the number of columns

    ## random permutation of Labels ~ traits
    y_col = 1#first column is randomLabel
    x_col = c(2:dim(pan2)[2])

    model<-as.formula(paste(colnames(pan2)[y_col], "~",
                            paste(colnames(pan2)[x_col],collapse = "+"),
                            sep = ""))


     gbm2<- gbm(model,
                   data=training2,
                   distribution="bernoulli",
                   n.trees=n.trees,
                   shrinkage=shrinkage,
                   interaction.depth=interaction.depth,
                   bag.fraction=0.50,
                   train.fraction=1,
                   n.minobsinnode=n.minobsinnode,
                   cv.folds=cv.folds,
                   keep.data=TRUE)

    #check performance using 5-fold cross-validation
    best.iter2 <- gbm.perf(gbm2,method="cv",plot.it=FALSE) #OOB method under predicts
    #   batsum2<-summary.gbm(gbm2,n.trees=best.iter,method=relative.influence,plotit=FALSE)
    best.iter2.list = c(best.iter2.list, best.iter2)
    ## LABEL
    ## predictions on the TRAINING SET
    output2<-predict(gbm2, newdata=training2, n.trees=best.iter2, type="response")
    output2<-cbind(output2,as.numeric(training2$randomLabel))

    # # training AUC for Bernoulli distributed responses
    auc2=colAUC(output2[,1],output2[,2])

    # Predictions on the TEST set
    output.test2<-predict(gbm2, newdata=test2, n.trees=best.iter2, type="response")
    output.test2<-cbind(output.test2,as.numeric(test2$randomLabel))
    # colnames(output.test2)<-c("output","label")
    # output.test2<-output.test2[order(-output.test2[,1]),]
    # plot(output.test)

    ## test AUC for Bernoulli distributed responses
    auctest2=colAUC(output.test2[,1],output.test2[,2])

    permutedAUC[i]<-auctest2
    print(auctest2)
    i=i+1
    print(i)#check where we are in bootstrap
  } else i=i
}
save(best.iter2.list, file = "best.iter2.list.presence.null.Rdata")
sum(is.na(permutedAUC)*1) #how many NAs
permutedAUC2<-na.omit(permutedAUC)
mean(permutedAUC2)
mean_AUC_presence_null = mean(permutedAUC2)
save(mean_AUC_presence_null, file= "mean_AUC_presence_null.Rdata")
load("mean_AUC_presence_null.Rdata")
sd(permutedAUC2)

#Stop the clock
(proc.time()-ptm)/60


```

##STOP HERE
```{r}
###### stop here for now ######


# predictions for all ticks
combined<-rbind(train,test)
output.comb<-predict(tickgbm, newdata=combined, n.trees=best.iter, type="response") 
output.comb<-cbind(output.comb,as.numeric(combined$GIDEON))
colnames(output.comb)<-c("output","label")
rownames(output.comb)<-combined$Species.name
output.comb<-output.comb[order(-output.comb[,1]),]

# testing hypotheses from data-mining
# need to update variable names
#1. do larvae with shorter tarsus I have greater host diversity (generalists)? YES
# plot(tick$L.Tarsus.I,tick$X.of.orders, 
#      xlab="Tarsus I length larvae", 
#      ylab="Orders infested")
# summary(lm(tick$L.Tarsus.I~tick$X.of.orders))

#2. are larvae with shorter tarsus I than expected for their body size the ones 
## who are generalists?
# plot(tick$residL.T.1,tick$X.of.orders, 
#      xlab="resid Tarsus I length", 
#      ylab="Orders infested")
# summary(lm(tick$L.Tarsus.I~tick$X.of.orders))

#3. produce a 2-panel figure that shows that the tick species that have shorter tarsus I lengths
# (for their body size; residual values below zero) also infest the greatest diversity of host species (# Orders)
# to do this, construct a dataframe of cols: species name, residual L.Tarsus.1, X.of.orders
# hyp3<-tick[,c(1,2,74,60)]
```







